Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2023-12-26 Manuscript_Nater_etal.qmd
Added code chunk for printing relevant stats for methods section

---
title: "Large-scale spatiotemporal variation in vital rates of an alpine bird"
format: html
author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public (also called citizen science, +ref), allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias (+ref). Consequently, many large-scale monitoring programmes are limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ ref). 
The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in >120 localities across the country (>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is noted, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups > 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) as well as bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively). <!-- I changed "owners" to "administrators", just because to me "public" implies it is not owned by anyone. But if this is wrong, please change it back -->

@@

In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different survey areas spanning 9 counties and 50 municipalities. Transects on which no willow ptarmigan were ever observed (i.e. species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.

@@

-   Introduction of integrated Distance sampling model (IDSM) from MS1

@@

@@

The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.

@@

@@

## Implications for management 
<!-- preliminary section title) -->

>>>>

---
title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
format: html
author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public (also called citizen science, +ref), allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias (+ref). Consequently, many large-scale monitoring programmes are limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ ref). The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is noted, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) as well as bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively). <!-- I changed "owners" to "administrators", just because to me "public" implies it is not owned by anyone. But if this is wrong, please change it back -->

@@

In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities. Transects on which no willow ptarmigan were ever observed (i.e. species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.

@@

@nilsen2023 recently developed an integrated distance sampling model (IDSM) which jointly analyses line transect and radio-telemetry data and applied it to willow ptarmigan in the Western part of Lierne municipality in Norway. The model consists of a deterministic population model with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival. Below, we describe our new extension of this model to include data from several as opposed to just one. For more detailed information on the single-site model, including tests of model performance, see @nilsen2023.

@@

For applying the ptarmigan IDSM across all 41 reporting districts we included an area index in all model parameters (@fig-model) and enabled sharing of information among areas by explicitly modelling spatial variation alongside shared temporal and residual variation in vital rates and detection parameters.

The spatially-explicit formulation of the two age-class population model can be written as:

$$
\begin{aligned}
  D_{juv, x, t+1} & = D_{ad, x, t+1} * R_{x, t+1} \\
  D_{ad, x, t+1} & = S_{x, t} * (D_{juv, x, t} + D_{ad, x, t})  
\end{aligned}
$$

Here, $D_{juv, x, t}$ and $D_{ad, x, t}$ are the densities of juvenile and adult ptarmigan in area $x$ in year $t$, respectively. Both juveniles and adult survive from year $t$ to $t+1$ with an area- ($x$) and year- ($t$) specific survival probability $S_{x, t}$, and survivors produce the next generation of juveniles according to an area- and year-specific recruitment rate ($R_{x,t}$).

@@

@@

```{r, echo = FALSE}
# Posterior summaries
post.sum <- readRDS("Results/PosteriorSummaries_byArea.rds")

# Printing function
printPostSum <- function(table, row = NULL, area = NULL){

  if(!missing(area)){
    row <- which(table$Area == area)
  }

  print.out <- list(
    area.name = table$Area[row],
    post.sum = paste0(
      round(table$Median[row], digits = 3), 
      " [",
      round(table$lCI[row], digits = 3),
      ", ",
      round(table$uCI[row], digits = 3),
      "]"
    )
  )
  return(print.out)
}

# Detection stats
#Det.hyper.mean
#Det.sd.spatial
#Det.sd.temporal
#Det.sd.residual
#Det.min
#Det.max

# Survival stats
#S.hyper.mean
#S.sd.spatial
#S.sd.temporal
#S.sd.residual
S.min <- printPostSum(table = post.sum$pSurv.sum, 
                      row = which(post.sum$pSurv.sum$Median == min(post.sum$pSurv.sum$Median)))
S.max <- printPostSum(table = post.sum$pSurv.sum,
                      row = which(post.sum$pSurv.sum$Median == max(post.sum$pSurv.sum$Median)))

# Recruitment stats
#R.hyper.mean
#R.sd.spatial
#R.sd.temporal
#R.sd.residual
R.min <- printPostSum(table = post.sum$rRep.sum, 
                      row = which(post.sum$rRep.sum$Median == min(post.sum$rRep.sum$Median)))
R.max <- printPostSum(table = post.sum$rRep.sum,
                      row = which(post.sum$rRep.sum$Median == max(post.sum$rRep.sum$Median)))

# Rodent effects stats
#betaR.hyper.mean
#betaR.sd.spatial
#betaR.sd.temporal
#betaR.sd.residual
betaR.min <- printPostSum(table = post.sum$betaR.sum, 
                      row = which(post.sum$betaR.sum$Median == min(post.sum$betaR.sum$Median)))
betaR.max <- printPostSum(table = post.sum$betaR.sum,
                      row = which(post.sum$betaR.sum$Median == max(post.sum$betaR.sum$Median)))
n.pos.betaR <- length(which(post.sum$betaR.sum$Median > 0))
n.neg.betaR <- length(which(post.sum$betaR.sum$Median < 0))

# Population density stats
popDens.sum <- subset(post.sum$popDens.sum, SummaryPeriod == "2018-2021")
dens.min <- printPostSum(table = popDens.sum, 
                      row = which(popDens.sum$Median == min(popDens.sum$Median)))
dens.max <- printPostSum(table = popDens.sum,
                      row = which(popDens.sum$Median == max(popDens.sum$Median)))

# Population growth rate stats
lambda.sum <- subset(post.sum$lambda.sum, SummaryPeriod == "2018-2021")
lambda.min <- printPostSum(table = lambda.sum, 
                      row = which(lambda.sum$Median == min(lambda.sum$Median)))
lambda.max <- printPostSum(table = lambda.sum,
                      row = which(lambda.sum$Median == max(lambda.sum$Median)))
n.pos.lambda <- length(which(lambda.sum$Median > 1))
n.neg.lambda <- length(which(lambda.sum$Median < 1))

```

@@

## Implications for management

<!-- preliminary section title) -->
